Harnessing generative AI to annotate the severity of all phenotypic abnormalities within the Human Phenotype Ontology

### 2024-05-23 docs/_tex/manuscript.tex
address NGS hypothesis comments

The Human Phenotype Ontology (HPO) has played a crucial role in
defining, diagnosing, prognosing, and treating human diseases by
providing a standardised framework for describing phenotypic
abnormalities. However, documentation on the severity of each phenotype
is currently very sparse, making systematic analyses and prioritisation
of results challenging. With \textgreater17500 phenotypic abnormalities
across \textgreater8600 rare diseases, manual curation of such
phenotypic annotations by experts would be exceedingly labour-intensive
and time-consuming. Leveraging advances in artificial intelligence, we
employed the OpenAI GPT-4 large language model (LLM) to systematically
annotate the severity of all phenotypic abnormalities in the HPO. First,
we benchmarked the generative LLM annotations against ground-truth
labels within the HPO (e.g.~phenotypes in the `Cancer' HPO branch were
annotating as causing cancer by GPT-4). True positive recall rates
across different annotations ranged from 89-100\% (mean=96\%), clearly
demonstrating the ability of GPT-4 to automate the curation process with
a high degree of fidelity. Using a novel approach, we developed a
severity scoring system that incorporates both the nature of the
phenotype outcome and the frequency of its occurrence. These severity
metrics will enable efforts to systematically prioritise which human
phenotypes are most detrimental to human health, and best targets for
therapeutic intervention.

@@

Comprehensive annotation of phenotypic abnormalities is invaluable for
defining, diagnosing, prognosing, and treating human disease. Since
2008, the Human Phenotype Ontology (HPO) has been instrumental to this,
by providing a standardised database for the description and analysis of
human phenotypes (Gargano et al., 2024; Köhler et al., 2021). Through
developing open community resources, the depth and breadth of the HPO
has continued to expand and there are now \textgreater17500 phenotypic
abnormalities across \textgreater8600 diseases, described. In recent
years, the HPO has expanded its disease annotations so that each HPO
term can have metadata including typical age of onset and frequency. In
addition, there are the \emph{Clinical modifier} and \emph{Clinical
course} subontologies, which contains terms to describe factors
including severity, triggers, mortality and progression, respectively.
Describing the severity-related attributes of a disease is crucial for
both research and clinical care of individuals with rare diseases. This
includes enhancing diagnostic capabilities, as well as prioritising and
guiding gene therapy trials.

To date, the HPO has largely relied on manual curation by experts
including clinicians, clinical geneticists and researchers. Although
this approach ensures the quality and accuracy of the ontology, it is
time-consuming and labour-intensive. As a result, less than 1\% of terms
within the HPO have metadata pertaining to their features such as time
course and severity. Artificial intelligence (AI) capabilities have
advanced considerably in recent years, presenting new opportunities to
integrate natural language processing technologies into assisting in the
curation process. Specifically, there have recently been considerable
advances in large language model (LLM) and their application to
biomedical problems, in some cases performing as well or better than
human clinicians on standardised medical exams and patient diagnosis
tasks (Bolton et al., 2024; Cheng et al., 2023; Gu et al., 2021; Labrak
et al., 2024; Luo et al., 2022; McDuff et al., 2023; O'Neil et al.,
2024; Shin et al., 2020; Singhal, Azizi, et al., 2023, 2023; Singhal,
Tu, et al., 2023; Van Veen et al., 2024; Zhang et al., 2023). Recent
work has demonstrated that the Generative Pre-trained Transformer 4
(GPT-4) foundation model (OpenAI et al., 2024), when combined with
strategic prompt engineering, can outperform even specialist LLMs that
are explicitly fine-tuned for biomedical tasks (Nori et al., 2023). In a
landmark achievement, GPT-4 was the first LLM to surpass a score of 90\%
in the United States Medical Licensing Examination (USML) (Nori et al.,
2023).

@@

793 phenotypes were annotated multiple times to evaluate the consistency
of responses generated by GPT-4. For a subset of phenotypes with known
expected annotations, true positive rates were calculated to assess
recall. Additionally, based on the clinical characteristics and their
occurrence, we have quantified the severity of each phenotype, providing
an example of how these clinical characteristic annotations can be used
to guide prioritisation of gene therapy trials. Ultimately, we hope that
our resource will be of utility to those working in rare diseases, as
well as the wider healthcare community.

@@

We employed the OpenAI GPT-4 model with Python to annotate 17502 terms
within the HPO (v2024-02-08) (Gargano et al., 2024; Köhler et al.,
2021). Our annotation framework was developed based on previously
defined criteria for classifying disease severity (Lazarin et al.,
2014). We sought to evaluate the impact of phenotypes on factors
including intellectual disability, death, impaired mobility, physical
malformations, blindness, sensory impairments, immunodeficiency, cancer,
reduced fertility, and congenital onset. Through prompt design we found
that the performance of GPT-4 improved when we incorporated a scale
associated with each effect and required a justification for each
response. For each effect, we asked about the likelihood of its
occurrence - whether it never, rarely, often, or always occurred.

@@

Less than 1\% of phenotype always directly resulted in death, such as
`Stillbirth', `Anencephaly' and `Bilateral lung agenesis'. Only 71
phenotypes were annotated as always causing death, while 9707 phenotypes
were annotated as often or rarely causing death. 7880 phenotypes were
annotated as never causing death. This included 34 unique forms of
syndactyly, a non-lethal condition that causes fused or webbed fingers
(occurring 1 in 1,200--15,000 live births). While not life-threatening
itself, syndactyly is a symptom of genetic disorders that can cause
life-threatening cardiovascular and neurodevelopmental defects, such as
Apert Syndrome (Garagnani \& Smith, 2013). This example highlights the
ability of GPT-4 to successfully distinguish between phenotypes that
directly cause lethality, and those that are often associated with
diseases that cause lethality.

@@

To assess annotation consistency, 793 we queried GPT-4 with a subset of
the HPO phenotypes multiple times. We employed two different metrics to
determine the \emph{consistency rate}. The first, less stringent metric,
defined consistency as the duplicate annotations being either `always'
and `often', or `never' and `rarely'. The second, more stringent metric,
required exact agreement in annotation occurrences, e.g.~`always' and
`always'. For the less stringent metric, duplicated phenotypes were
annotated consistently at a rate of at least 80\%, and for the more
stringent metric, the lowest consistency rate was 57\% for congenital
onset. An example of where annotations were inconsistent was for the HPO
term `Acute leukaemia'. One time, GPT-4 annotated it as often causing
impaired mobility, giving the justification that `weakness and fatigue
from leukaemia and its treatment can impair mobility'. The other time,
GPT-4 annotated it as rarely causing impaired mobility, giving the
justification that `acute leukaemia rarely impairs mobility directly'.
Despite specifying in the prompt for GPT-4 not to take into
consideration indirect effects, this is an example of where it failed to
do so.

@@

This high level of recall underscores the robustness of our annotations
and the reliability of the HPO framework in capturing clinically
relevant phenotypic information. However, we acknowledge that the number
of testable true positive phenotypes for some of these categories are
low, especially `blindness' for which there is only 1 phenotype in the
HPO (after excluding terms pertaining to colour or night blindness).
Furthermore, some of the true positive phenotypes are lexically similar
to the name of the clinical characteristic itself. In these cases,
annotating `Severe intellectual disability' as always causing
intellectual disability is a relatively trivial task. Nevertheless, even
these scenarios provide a clear and interpretable benchmark for the
model's performance. In addition, were numerous phenotypes with
lexically non-obvious relationships to the clinical characteristic that
were annotated correctly by GPT-4. For example, `Molar tooth sign on
MRI' was correctly annotated as causing intellectual disability.

@@

Quantifying the severity of phenotypes can have important implications
for diagnosis, prognosis, and treatment, for instance, guiding
prioritisation of gene therapy trials for the most severe phenotypes.
First, we created a dictionary to map each phenotype outcome
(e.g.~blindness) and its response (always, often, rarely, never) to
numeric values from 0-3. Then, the phenotype outcome values were
multiplied by their associated response values. Importantly, the values
reflected the severity of each outcome based on both the outcome itself
and the frequency of the response. For instance, a phenotype always
causing death would have a higher multiplied value than a phenotype
rarely causing reduced fertility (see Table~\ref{tbl-metric-weights}).
Next, we computed an average score for each phenotype by aggregating the
multiplied values across all phenotype outcomes and then calculating the
mean. This was then normalised by the theoretical maximum severity
score, so that all phenotypes were on a 0-100 severity scale (where 100
is the most severe phenotype possible). This average normalised score
represents the overall severity of the phenotype based on the severity
of its individual outcomes.

@@

\caption{\label{fig-top-phenos}Quantifying the severity of HPO phenotype
annotations highlights the most impactful conditions. Heatmap of 10
represetantive phenotypes from each severity class (Profound, Severe,
Moderate, Mild) stratified by whether the phenotypes are often/always
congenital (\textbf{a}-\textbf{b}) or rarely/never congenital
(\textbf{c}-\textbf{d}). Continuous severity scores are shown as bars
(\textbf{b},\textbf{d}) and were calculated by multiplying the numeric
values assigned to each clinical characteristic according to
Table~\ref{tbl-metric-weights}. The average normalised score,
representing overall phenotype severity on a 0-100 scale, was calculated
by aggregating the multiplied values and normalising by the theoretical
maximum severity score. The x-axes show each of the clinical
characteristics.}

@@

@@

Here, we introduce a novel framework to leverage the current
best-in-class LLM, GPT-4 (OpenAI et al., 2024), to systematically
annotate the severity of 17502 phenotypic abnormalities within the HPO.
By employing advanced AI capabilities, we have demonstrated the
feasibility of automating this process, significantly enhancing
efficiency without substantially compromising accuracy. Our validation
approach yielded a high true positive rate exceeding 88\% across the
phenotypes tested. Furthermore, our approach can be readily adapted and
scaled to accommodate the growing volume of phenotypic data.

@@

Our findings highlight the potential of this next generation of natural
language processing technologies in significantly contributing to the
automation and refinement of data curation in biomedical research. These
results have a large number of useful real-world applications, such as
prioritising gene therapy candidates (Murphy et al., 2023) and guiding
clinical decision-making in rare diseases. As the HPO continues to grow
over time, this approach can easily be repeated to keep pace with the
rapidly evolving biomedical landscape. Furthermore, this approach can be
extended to produce different sets of annotations or be used with any
other ontology. Additional use cases include gathering data on the
prevalence of each phenotype to approximate their social and financial
costs.

@@

@@

We wrote a Python script to iteratively query GPT-4 via the OpenAI
application programming interface (API). The ultimately yielded
consistently formatted annotations for 17502 terms within the HPO. Our
annotation framework was developed based on previously defined criteria
for classifying disease severity (Lazarin et al., 2014). We sought to
evaluate whether each phenotype directly caused a given severity-related
clinical characteristic, including: intellectual disability, death,
impaired mobility, physical malformations, blindness, sensory
impairments, immunodeficiency, cancer, reduced fertility, and/or had a
congenital onset. Through prompt engineering we found that the
performance of GPT-4 improved when we incorporated a scale associated
with each effect and required a justification for each response. We
asked how frequently the given phenotype directly causes each clinical
characteristic - whether it never, rarely, often, or always occurred.
Furthermore, our prompt design revealed that the optimal trade-off
between the number of phenotypes and performance (in terms of producing
the desired annotations, and adhering to the formatting requirements)
was achieved when inputting no more than two or three phenotypes per
prompt. An example prompt can be seen in Figure~\ref{fig-occurrence}.
Only 2 phenotypes were included per prompt in order to 1) avoid
exceeding per-query token limits, and 2) prevent the breakdown of GPT-4
performance due to long-form text input, which is presently a known
limitation common to many LLMs including GPT-4 (Wei et al., 2024).

@@

>>>>

There are thousands of human phenotypes which are linked to genetic
variation. These range from the benign (white eyelashes) to the deadly
(respiratory failure). The Human Phenotype Ontology has categorised all
human phenotypic variation into an unified framework that defines the
relationships between them (e.g.~missing arms and missing legs are both
abnormalities of the limb). This has made it possible to perform
phenome-wide analyses, e.g.~to prioritise which make the best candidates
for gene therapies. However, there is currently limited metadata
describing the effects / severity of these phenotypes. With
\textgreater17500 phenotypic abnormalities across \textgreater8600 rare
diseases, manual curation of such phenotypic annotations by experts
would be exceedingly labour-intensive and time-consuming. Leveraging
advances in artificial intelligence, we employed the OpenAI GPT-4 large
language model (LLM) to systematically annotate the severity of all
phenotypic abnormalities in the HPO. First, we benchmarked the
generative LLM annotations against ground-truth labels within the HPO
(e.g.~phenotypes in the `Cancer' HPO branch were annotating as causing
cancer by GPT-4). True positive recall rates across different
annotations ranged from 89-100\% (mean=96\%), clearly demonstrating the
ability of GPT-4 to automate the curation process with a high degree of
fidelity. Using a novel approach, we developed a severity scoring system
that incorporates both the nature of the phenotype outcome and the
frequency of its occurrence. These severity metrics will enable efforts
to systematically prioritise which human phenotypes are most detrimental
to human health, and best targets for therapeutic intervention.

@@

Ontologies provide a common language with which to communicate concepts.
In medicine, ontologies for phenotypic abnormalities are invaluable for
defining, diagnosing, prognosing, and treating human disease. Since
2008, the Human Phenotype Ontology (HPO) has been instrumental in
healthcare and biomedical research by providing a framework for
comprehensively describing human phenotypes and the relationships
between them (Gargano et al., 2024; Köhler et al., 2021). By expanding
its depth and breadth over time, the HPO now contains \textgreater17500
phenotypic abnormalities across \textgreater8600 diseases. Some HPO
phenotypes also contain metadata annotations such typical age of onset,
frequency, triggers, time course, mortality rate and typical severity.
Describing the severity-related attributes of a disease is crucial for
both research and clinical care of individuals with rare diseases. When
researchers or clinicians are presented with phenotypes that fall
outside of their expertise, resources to quickly and reliably retrieve
summaries with additional relevant information about these phenotypes
are essential. In the clinic, this can help in reaching a differential
diagnosis or prioritising the treatment of some phenotypes over others.
In research, this information is useful for prioritising targets for
causal disease mechanisms, performing large-scale analyses of phenotypic
data, and guiding funding agencies when assessing the potential impact
and need for research in a given disease area. To date, the HPO has
largely relied on manual curation by domain experts. While this approach
can improve annotation quality and accuracy, it is both time-consuming
and labour-intensive. As a result, less than 1\% of terms within the HPO
contain metadata such as time course and severity.

Artificial intelligence (AI) capabilities have advanced considerably in
recent years, presenting new opportunities to integrate natural language
processing technologies into assisting in the curation process.
Specifically, there have recently been considerable advances in large
language model (LLM) and their application to biomedical problems, in
some cases performing as well or better than human clinicians on
standardised medical exams and patient diagnosis tasks (Bolton et al.,
2024; Cheng et al., 2023; Gu et al., 2021; Labrak et al., 2024; Luo et
al., 2022; McDuff et al., 2023; O'Neil et al., 2024; Shin et al., 2020;
Singhal, Azizi, et al., 2023, 2023; Singhal, Tu, et al., 2023; Van Veen
et al., 2024; Zhang et al., 2023). Recent work has demonstrated that the
Generative Pre-trained Transformer 4 (GPT-4) foundation model (OpenAI et
al., 2024), when combined with strategic prompt engineering, can
outperform even specialist LLMs that are explicitly fine-tuned for
biomedical tasks (Nori et al., 2023). In a landmark achievement, GPT-4
was the first LLM to surpass a score of 90\% in the United States
Medical Licensing Examination (USML) (Nori et al., 2023).

@@

To evaluate the consistency of responses generated by GPT-4 793
phenotypes were annotated multiple times. For a subset of phenotypes
with known expected annotations, true positive rates were calculated to
assess recall. Additionally, based on the clinical characteristics and
their occurrence, we have quantified the severity of each phenotype,
providing an example of how these clinical characteristic annotations
can be used to guide prioritisation of gene therapy trials. Ultimately,
we hope that our resource will be of utility to those working in rare
diseases, as well as the wider healthcare community.

@@

We employed the OpenAI GPT-4 model with Python to annotate 17502 terms
within the HPO (v2024-02-08) (Gargano et al., 2024; Köhler et al.,
2021). Our annotation framework was developed based on previously
defined criteria for classifying disease severity (Lazarin et al.,
2014). We sought to evaluate the impact of phenotypes on factors
including intellectual disability, death, impaired mobility, physical
malformations, blindness, sensory impairments, immunodeficiency, cancer,
reduced fertility, and congenital onset. Through prompt design we found
that the performance of GPT-4 improved when we incorporated a scale
associated with each effect and required a justification for each
response. For each effect, we asked about the frequency of its
occurrence - whether it never, rarely, often, or always occurred.
Framing the queries in this way served two purposes. First, this helped
to constrain the responses of GPT-4 to a specific range of values,
making answers more consistent and amenable to downstream data analysis.
Second, it served to overcome one of the main limitations noted by
Lazarin et al. (2014) as they did not collect information on how the
frequency of each disease affected their decision making when generating
severity annotations.

@@

Less than 1\% of phenotypes always directly resulted in death (n=71),
such as `Stillbirth', `Anencephaly' and `Bilateral lung agenesis'.
Meanwhile, 9707 phenotypes were annotated as often or rarely causing
death. 7880 phenotypes were annotated as never causing death. Examples
of phenotypes that never cause death included 34 unique forms of
syndactyly, a non-lethal condition that causes fused or webbed fingers
(occurring 1 in 1,200--15,000 live births). While not life-threatening
itself, syndactyly is a symptom of genetic disorders that can cause
life-threatening cardiovascular and neurodevelopmental defects, such as
Apert Syndrome (Garagnani \& Smith, 2013). This example highlights the
ability of GPT-4 to successfully distinguish between phenotypes that
directly cause lethality, and those that are often associated with
diseases that cause lethality.

@@

To assess annotation consistency, we queried GPT-4 with a subset of the
HPO phenotypes multiple times (n=793 unique phenotypes). We employed two
different metrics to determine the \emph{consistency rate}. The first,
less stringent metric, defined consistency as the duplicate annotations
being either `always' and `often', or `never' and `rarely'. The second,
more stringent metric, required exact agreement in annotation
occurrences, e.g.~`always' and `always'. For the less stringent metric,
duplicated phenotypes were annotated consistently at a rate of at least
80\%, and for the more stringent metric, the lowest consistency rate was
57\% for congenital onset. An example of where annotations were
inconsistent was for the HPO term `Acute leukaemia'. One time, GPT-4
annotated it as often causing impaired mobility, giving the
justification that `weakness and fatigue from leukaemia and its
treatment can impair mobility'. The other time, GPT-4 annotated it as
rarely causing impaired mobility, giving the justification that `acute
leukaemia rarely impairs mobility directly'. Despite specifying in the
prompt for GPT-4 not to take into consideration indirect effects, this
is an example of where it failed to do so.

@@

This high level of recall underscores the robustness of our annotations
and the reliability of the HPO framework in capturing clinically
relevant phenotypic information. However, we acknowledge that the number
of testable true positive phenotypes for some of these categories are
low, especially `blindness' for which there is only 1 phenotype in the
HPO (after excluding terms pertaining to colour or night blindness).
Furthermore, some of the true positive phenotypes are lexically similar
to the name of the clinical characteristic itself. In these cases,
annotating `Severe intellectual disability' as always causing
intellectual disability is a relatively trivial task. Nevertheless, even
these scenarios provide a clear and interpretable benchmark for the
model's performance. In addition, were numerous phenotypes with
lexically non-obvious relationships to the clinical characteristic that
were annotated correctly by GPT-4. For example, `Molar tooth sign on
MRI' (a neurodevelopmental pathology observed in radiological scans) was
correctly annotated as causing intellectual disability.

@@

While individual annotations are informative, we wanted to be able to
distil the severity of each phenotype into a single score. Quantifying
the overall severity of phenotypes can have important implications for
diagnosis, prognosis, and treatment. It may also guide the
prioritisation of gene therapy trials for phenotypes with the most
severe outcomes and thus the most urgent need. Importantly, the values
reflected the severity of each outcome based on both the type of outcome
itself and its frequency within a particular phenotype. For instance, a
phenotype always causing death would have a higher multiplied value than
a phenotype often causing reduced fertility (see
Table~\ref{tbl-metric-weights}). First, we created a dictionary to map
each phenotype outcome (e.g.~blindness) and its frequency (always,
often, rarely, never) to numeric values from 0-3. Then, the phenotype
outcome values were multiplied by phenotype outcome weights. Next, we
computed an average score for each phenotype by aggregating the
multiplied values across all phenotype outcomes and then calculating the
mean. This was then normalised by the theoretical maximum severity
score, so that all phenotypes were on a 0-100 severity scale (where 100
is the most severe phenotype possible). This average normalised score
represents the overall severity of the phenotype based on the severity
of its individual outcomes.

@@

\caption{\label{fig-top-phenos}Quantifying the severity of HPO phenotype
annotations highlights the most impactful conditions. Heatmap of 10
represetantive phenotypes from each severity class (Profound, Severe,
Moderate, Mild) stratified by whether the phenotypes are often/always
congenital (\textbf{a}-\textbf{b}) or rarely/never congenital
(\textbf{c}-\textbf{d}). Continuous severity scores are shown as bars
(\textbf{b},\textbf{d}) and were calculated by multiplying the numeric
values assigned to each clinical characteristic according to
Table~\ref{tbl-metric-weights}. The average normalised score,
representing overall phenotype severity on a 0-100 scale, was calculated
by aggregating the multiplied values and normalising by the theoretical
maximum severity score. The x-axes show each of the clinical
characteristics. All data for this figure, as well as justifications for
each annotation, can be found in Table~\ref{tbl-annotations}.}

@@

Phenotype severity annotations have utility across a wide variety of
applications in both the clinic and research. In clinical settings,
severity annotations can be used to prioritise the treatment of some
phenotypes over others in patients with complex presentations, avoid
administering contraindicated drugs, and prognosing potential health
outcomes. In research settings, severity annotations can be used to
identify phenotypes that have a large impact on patient outcomes and yet
are currently understudied. They may also be used to help design new
experiments and studies, or even provide new insights into the
underlying aetiology of the disease by making expert-level summaries
more immediately accessible to the wider research community.

@@

Here, we introduce a novel framework to leverage the current
best-in-class LLM, GPT-4 (OpenAI et al., 2024), to systematically
annotate the severity of 17502 phenotypic abnormalities within the HPO.
By employing advanced AI capabilities, we have demonstrated the
feasibility of automating this process, significantly enhancing
efficiency without substantially compromising accuracy. Our validation
approach yielded a high true positive rate exceeding 88\% across the
phenotypes tested. Furthermore, our approach can be readily adapted and
scaled to accommodate the growing volume of phenotypic data. In total,
the entire study cost \$296.27 in queries to the OpenAI API. While we do
not have a direct comparison, this likely represents a extremely small
fraction of the total costs of such a study if performed manually by
human experts charging at an hourly rate. Even if all human annotations
were provided on a volunteer basis, this would still require hundreds if
not thousands of hours of cumulative manual human labour. Using our
approach, severity annotations for the entire HPO can be generated
automatically in a matter of several hours.

@@

Our findings highlight the potential of this next generation of natural
language processing technologies in significantly contributing to the
automation and refinement of data curation in biomedical research. These
results have a large number of useful real-world applications, such as
prioritising gene therapy candidates (Murphy et al., 2023) and guiding
clinical decision-making in rare diseases. It may also be used as tool
to help inform policy decisions and funding allocation by healthcare or
governmental institutions. This of course would need to be in
consultation with subject matter medical experts, patients, advocates
and biomedical ethicists before reaching a final decision. Nevertheless,
access to succinct, interpretable, and semi-quantitative severity
annotations may encourage key decision makers with limited time to
review individual proposals to pay heed to phenotypes and diseases that
would otherwise be overlooked. As the HPO and the broader literature
continue to grow over time, our automated AI-based approach can easily
be repeated to keep pace with the rapidly evolving biomedical landscape.
Furthermore, it can be extended to produce different sets of annotations
or be used with any other ontology. Additional use cases include
gathering data on the prevalence of each phenotype to approximate their
social and financial costs.

@@

Another limitation with the present dataset is that phenotypes
themselves can manifest with different degrees of severity, in the sense
that they are more pronounced or intense. For example, sensitivity to
light could range from a mild inconvenience to a severe disability that
prevents the individual from leaving their home during the day. The
effect of onset (beyond congenital vs.~non-congenital) and time course
(acute, slowly progression, relapse-remitting) were also not explicitly
considered. Finally, we did not ask GPT-4 to consider phenotypes as they
present within particular diseases. For example, while the phenotype
`Hypertension' may be mild to moderate in the general population and not
present until middle-age, it can also present early in life as very
severe in the context of a rare genetic disorder such as Liddle
syndrome. Future work could explore these nuances in more detail.

@@

We wrote a Python script to iteratively query GPT-4 via the OpenAI
application programming interface (API). The ultimately yielded
consistently formatted annotations for 17502 terms within the HPO. Our
annotation framework was developed based on previously defined criteria
for classifying disease severity (Lazarin et al., 2014). We sought to
evaluate whether each phenotype directly caused a given severity-related
clinical characteristic, including: intellectual disability, death,
impaired mobility, physical malformations, blindness, sensory
impairments, immunodeficiency, cancer, reduced fertility, and/or had a
congenital onset. Through prompt engineering we found that the
performance of GPT-4 improved when we incorporated a scale associated
with each effect and required a justification for each response. We
asked how frequently the given phenotype directly causes each clinical
characteristic - whether it never, rarely, often, or always occurred.
This design helps to constrain the potential responses of GPT-4 and thus
make it more amenable to machine-readable post-processing. It also
serves to address one of its key limitations from the Lazarin et al.
(2014) survey, namely the lack information on how clinical
characteristic frequency affected the clinicians' severity annotations.
Here, we can instead use the frequency values to generate more precise
annotations and downstream severity ranking scores.

Furthermore, our prompt design revealed that the optimal trade-off
between the number of phenotypes and performance (in terms of producing
the desired annotations, and adhering to the formatting requirements)
was achieved when inputting no more than two or three phenotypes per
prompt. An example prompt can be seen in Figure~\ref{fig-occurrence}.
Thus, only two phenotypes were included per prompt in order to 1) avoid
exceeding per-query token limits, and 2) prevent the breakdown of GPT-4
performance due to long-form text input, which is presently a known
limitation common to many LLMs including GPT-4 (Wei et al., 2024).

@@

\begin{table}

\caption{\label{tbl-annotations}Table of GPT-4 annotations for all Human
Phenotype Ontology (HPO) phenotypes in Figure~\ref{fig-top-phenos}. For
each phenotype, this includes the name of the phenotype (`hpo\_name'),
the ID of the phenotype (`hpo\_id'), the frequency of each annotation
(always, often, rarely, never), and the justification for each
annotation ('\ldots\_justification'). These results can also be
downloaded programmatically using the R function
\texttt{HPOExplorer::gpt\_annot\_check}.}

\href{https://github.com/neurogenomics/gpt_hpo_annotations/raw/master/data/top_phenos_annotations.csv.gz}{\textbf{Top
phenotype annotations table}}

\end{table}%
